---
title: "Using and Improving Decision Trees in R"
subtitle: "Orange County R User Group"
author: "Robert Mohr"
date: "January 23, 2018"
fontsize: 6
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r set-options, echo=FALSE, cache=FALSE}

```
## Packages and References

There are a couple of standard decision tree packages, used by a couple of prominent modeling books:

- **tree**, *An Introduction to Statistical Learning* by James, Witten, Hastie, and Tibshirani
- **rpart**, *Applied Predictive Modeling* by Kuhn and Johnson

The visualizations supported by these packages are based on R's base graphics system.  We are developing a script, gridTree.R, that leverages the more powerful graphics system of the **grid** package.

Today, we will focus on **tree** and gridTree.R.

## **tree**: Fitting the data

Fitting is done with one call to tree().  tree() uses standard formula notation:

```{r cars, echo = TRUE}
library(tree)

tree.iris <- tree(Species~., data=iris)
```

This works for both classification and regression trees.

## Exploring the tree object: raw
```{r, echo=T}
tree.iris
```

## Exploring the tree object: str()
```{r, echo=T}
str(tree.iris, give.attr = F)
```

## Exploring the tree object: $frame
```{r, echo=T, size="tiny"}
str(tree.iris$frame)
```

## **tree**: Plotting the results

Plotting is straightforward, thanks to methods that **tree** provides for plot() and text() for objects of class tree.

```{r, echo=T}
# Calls to inspect the tree methods
# methods(plot)
# getS3method("plot", "tree")
# methods(text)
# getS3method("text", "tree")

# Two calls to render everything (results on next slide)
# plot(tree.iris)
# text(tree.iris)
```

## A **tree** plot of iris

```{r}
plot(tree.iris)
text(tree.iris)
```

## Opportunities for improvement: orientation

Decision trees are traditionally drawn top-down.  This has a couple of disadvantages:

- The major axis of text is horizontal rather than vertical, so overlap is inevitable.
- Labeling the branching criteria (< x, > y) usually involves dropping one label, which is ambiguous, or spreading labels out, which is inefficient.

Creating a left-right oriented decision tree would be generally easier to read.

## Opportunities for improvement: color encoding

Traditional decision trees make use of almost no data-encoding visual attributes.

For example, nodes could be color-encoded to represent:

-Classifications, as a redundancy to direct labelling, or in place of it
-Diagnostic measures, such as node purity.  Keep in mind this is effective for rough distinctions only.

## A very brief introduction to **grid**

- **grid** is a powerful graphics package that underlies **ggplot2** and **lattice** graphics
- **grid** can be used to modify these graphics or create entirely new ones
- Part of **grid**'s power comes from trees of *viewport* objects

For more detail, see *R Graphics*
by Paul Murrell (the creator of **grid**).

## A very brief introduction to viewports

- A viewport is a rectangular drawing region
- Viewports have a *geometric context* (coordinate systems, size) and a *graphical context* (graphical parameters like fontsize and colors)
- Viewports have a special coordinate system known as *normalized parent coordinates* (npc) with (0,0) at the lower left corner and (1,1) in the upper right corner
- Viewports can have children that inherit their properties

## Plotting with gridTree

gridTree.R takes advantage of the fact that both decision trees and viewports have a tree structure:

- grid.tree() accepts an object of class tree, extracts the data frame, and calls grid.grow() for the first tree node
- grid.grow() draws a part of the tree by calling grid.branch() for a split or grid.leaf() for a terminal node
- grid.grow() recursively calls itself for the children of the current node
- The process naturally terminates when there are no more nodes

## A gridTree rendering of iris

```{r}
source('~/R/gridTree/gridTree.R')
```

```{r}
library(grid)
# source('~/R/gridTree/gridTree.R')
grid.tree(tree.iris)
```

## Desirable features of the gridTree

- Can be easily read left-to-right
- No overlap of text with the tree
- Variables and their bounds are closely grouped and aligned
- Terminal nodes can be color-coded with supporting data

## Future work on gridTree

- Implementation of colored nodes
- Checks and evasions for colliding branches
- Checks and evasions for colliding text
- Implementation of graphical objects (grobs)

Repository on github: https://github.com/mohrsignal/gridTree

Email mohrsignal@gmail.com with questions.

Thanks for your time and attention!