--- 
title: "ggplot2: An Introduction"
author: "Arun Rangarajan"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
#output: bookdown::gitbook
#always_allow_html: yes
#documentclass: book
#bibliography: [learning_more.bib]
#biblio-style: apalike
#link-citations: yes
#github-repo: rstudio/bookdown-demo
#description: "Short intro to ggplot2."
---
# Background
* R has different plotting systems:
     + base
     + grid
     + lattice
     + ggplot2
* ggplot2 - based on the good parts of other systems
* gg = Grammar of Graphics
* Based on Leland Wilkinson’s “The Grammar of Graphics”
* Developed in R by Hadley Wickham
* This presentation is based on [Wickham, 2016] and [Kabacoff, 2015]
* (with very little "original" content :-))

# What is special about ggplot2?
* Deep underlying grammar (just like language grammar)
* Based on a simple set of core principles
* Good defaults 
* Lots of options for customization
* Takes care of a lot of details (like legends, axes grids)
* Plots are made iteratively
* Easy to learn

# Parts of a ggplot
* Data
* Aesthetics (aes)
     + Color
     + Shape
     + Size
* Geometric Objects (geoms)
     + Points
     + Lines
     + Polygons
* Statistical Transformations (stats)
     + Histograms
     + Boxplots
* Legends and Axes
* Coordinate system (coord)
     + Cartesian
     + Polar
* Faceting (aka latticing/trellising)
* Theme
     + Background color
     + Font size

# Installation and Loading
* ggplot2 is not shipped with base R, so:
```{r eval=FALSE}
install.packages("ggplot2")
```
* Load the library before use:
```{r eval=FALSE}
library(ggplot2)
```

# Data Set to Explore
* `ggplot2` requires the data be provided in a data frame
* Example data frame to explore: `mtcars`
```{r}
dim(mtcars)
head(mtcars)
```

# Our first ggplot
* ggplot2 uses the `+` symbol to chain functions
```{r}
library(ggplot2)
ggplot(data=mtcars, aes(x=wt, y=mpg)) + 
  geom_point()
```

# Parts of our first ggplot
```{r eval=FALSE}
ggplot(data=mtcars, aes(x=wt, y=mpg)) + 
  geom_point()
```
* `ggplot()` function initializes the plot
* Data frame specified in `data` param
* Variables specified in `aes` (aesthetics) param
     + x-axis is `wt`
     + y-axis is `mpg`
* `geom` specifies the visual output (`geom_point()` here 
   produces scatter plot)

# Adding title and axes labels
```{r}
ggplot(data=mtcars, aes(x=wt, y=mpg)) + 
  geom_point() + 
  labs(title="Auto Data", x="Weight", y="Miles per Gallon")
```

# Adding colors to points
* Color the points based on `cyl` column
```{r}
ggplot(data=mtcars, aes(x=wt, y=mpg)) + 
  geom_point(aes(color=cyl)) + 
  labs(title="Auto Data", x="Weight", y="Miles per Gallon")
```

* `cyl` is a numeric field, so continuous colors

# Adding colors to points (cont...)
* Convert `cyl` to factor
```{r}
ggplot(data=mtcars, aes(x=wt, y=mpg)) + 
  geom_point(aes(color=as.factor(cyl))) + 
  labs(title="Auto Data", x="Weight", y="Miles per Gallon", 
       color="Cylinders")
```

# Changing the shapes
```{r}
ggplot(data=mtcars, aes(x=wt, y=mpg)) + 
  geom_point(aes(color=as.factor(cyl), shape=as.factor(cyl))) + 
  labs(title="Auto Data", x="Weight", y="Miles per Gallon", 
       color="Cylinders")
```

# Fix the legend
```{r}
ggplot(data=mtcars, aes(x=wt, y=mpg)) + 
  geom_point(aes(color=as.factor(cyl), shape=as.factor(cyl))) + 
  labs(title="Auto Data", x="Weight", y="Miles per Gallon", 
       color="Cylinders", shape="Cylinders")
```

# Increase the point size
* Do this outside `aes`
```{r}
ggplot(data=mtcars, aes(x=wt, y=mpg)) + 
  geom_point(aes(color=as.factor(cyl), shape=as.factor(cyl)), size=2) + 
  labs(title="Auto Data", x="Weight", y="Miles per Gallon", 
       color="Cylinders", shape="Cylinders")
```

# Adding a loess smoother
```{r}
ggplot(data=mtcars, aes(x=wt, y=mpg)) + 
  geom_point(aes(color=as.factor(cyl), shape=as.factor(cyl)), size=2) + 
  geom_smooth() +
  labs(title="Auto Data", x="Weight", y="Miles per Gallon", 
       color="Cylinders", shape="Cylinders")
```

# Remove the confidence interval
```{r}
ggplot(data=mtcars, aes(x=wt, y=mpg)) + 
  geom_point(aes(color=as.factor(cyl), shape=as.factor(cyl)), size=2) + 
  geom_smooth(se=FALSE) +
  labs(title="Auto Data", x="Weight", y="Miles per Gallon", 
       color="Cylinders", shape="Cylinders")
```

# Add a linear fit
```{r}
ggplot(data=mtcars, aes(x=wt, y=mpg)) + 
  geom_point(aes(color=as.factor(cyl), shape=as.factor(cyl)), size=2) + 
  geom_smooth(se=FALSE) + 
  geom_smooth(method="lm", se=FALSE, color="purple", lty=2) +
  labs(title="Auto Data", x="Weight", y="Miles per Gallon", 
       color="Cylinders", shape="Cylinders")
```

# One fit per group
```{r}
ggplot(data=mtcars, aes(x=wt, y=mpg)) + 
  geom_point(aes(color=as.factor(cyl), shape=as.factor(cyl)), size=2) + 
  geom_smooth(se=FALSE, aes(group=as.factor(cyl), 
                            color=as.factor(cyl))) + 
  geom_smooth(method="lm", se=FALSE, lty=2, 
              aes(group=as.factor(cyl), 
                  color=as.factor(cyl))) +
  labs(title="Auto Data", x="Weight", y="Miles per Gallon", 
       color="Cylinders", shape="Cylinders")
```

# Separate the groups with faceting
```{r}
ggplot(data=mtcars, aes(x=wt, y=mpg)) + 
  geom_point(aes(color=as.factor(cyl), shape=as.factor(cyl)), size=2) + 
  geom_smooth(se=FALSE, aes(group=as.factor(cyl))) + 
  geom_smooth(method="lm", se=FALSE, color="purple", lty=2, 
              aes(group=as.factor(cyl))) +
  labs(title="Auto Data", x="Weight", y="Miles per Gallon") + 
  # remove the legend
  theme(legend.position="none") + 
  facet_wrap(~cyl)
```

# More Faceting
```{r}
mtcars$am <- factor(mtcars$am, levels=c(0, 1), labels=c("Auto", "Manual"))
mtcars$vs <- factor(mtcars$vs, levels=c(0, 1), labels=c("V-Engine", "Straight Engine"))
mtcars$cyl <- factor(mtcars$cyl)
ggplot(mtcars, aes(x=hp, y=mpg, color=cyl, shape=cyl)) + 
  geom_point(size=2) + 
  facet_grid(am ~ vs) + 
  labs(title="Auto Data by Engine and Transmission Type", x="Horsepower", y="mpg")
```

# Boxplots
```{r}
ggplot(mtcars) + 
  geom_boxplot(aes(x=as.factor(cyl), y=mpg, fill=as.factor(cyl))) + 
  labs(x="Cylinders") + 
  theme(legend.position="none")
```

# Violin Plots
```{r}
ggplot(mtcars) + 
  geom_violin(aes(x=as.factor(cyl), y=mpg, fill=as.factor(cyl))) + 
  labs(x="Cylinders") + 
  theme(legend.position="none")
```

# Combine boxplot with violin plot
```{r}
ggplot(mtcars) + 
  geom_violin(aes(x=as.factor(cyl), y=mpg), fill="lightblue") + 
  geom_boxplot(aes(x=as.factor(cyl), y=mpg), fill="lightgreen", width=0.1) + 
  labs(x="Cylinders") + 
  theme(legend.position="none")
```

# Histogram
```{r}
ggplot(mtcars) + 
  geom_histogram(aes(mpg), binwidth=1)
```

# Bar Plots
```{r}
library(car)
data(Salaries, package="car")
head(Salaries, 2)
ggplot(Salaries, aes(x=rank, fill=sex)) + 
  geom_bar(position="stack")
```

# Filled Bar Plots
```{r}
ggplot(Salaries, aes(x=rank, fill=sex)) + 
  geom_bar(position="fill") + 
  # change the default y label 'count' to 'proportion'
  labs(y="proportion")
```

# Changing the Y-axis breaks
```{r}
ggplot(Salaries, aes(x=rank, fill=sex)) + 
  geom_bar(position="fill", alpha=0.7) + 
  labs(y="proportion") + 
  scale_y_continuous(breaks=seq(0, 1, 0.1))
```

# Manual colors
```{r}
ggplot(Salaries, aes(x=rank, fill=sex)) + 
  geom_bar(position="fill", alpha=0.7) + 
  labs(y="proportion") + 
  scale_y_continuous(breaks=seq(0, 1, 0.1)) + 
  scale_fill_manual(values = c("brown", "orange"))
```

# Black and White Theme
```{r}
ggplot(Salaries, aes(x=rank, fill=sex)) + 
  geom_bar(position="fill", alpha=0.7) + 
  labs(y="proportion") + 
  scale_y_continuous(breaks=seq(0, 1, 0.1)) + 
  scale_fill_manual(values = c("brown", "orange")) + 
  theme_bw()
```

# Make ggplot interactive with plotly
```{r message=FALSE}
library(plotly)
p <- ggplot(data=mtcars, aes(x=wt, y=mpg)) + 
  geom_point(aes(color=as.factor(cyl), shape=as.factor(cyl)), 
             size=2) + 
  labs(title="Auto Data", x="Weight", y="Miles per Gallon", 
       color="Cylinders", shape="Cylinders")
ggplotly(p)
```

# Change the hover text
```{r message=FALSE}
library(plotly)
mtcars$rowName <- row.names(mtcars)
p <- ggplot(data=mtcars, aes(x=wt, y=mpg)) + 
  geom_point(aes(color=as.factor(cyl), 
                 shape=as.factor(cyl), 
                 text=paste(rowName)), 
             size=3) + 
  labs(title="Auto Data", x="Weight", y="Miles per Gallon", 
       color="Cylinders", shape="Cylinders")
ggplotly(p)
```

# Learning more {-}

* Hadley Wickham, _ggplot2: Elegant Graphics for Data Analysis_, 
  2/e, Springer, 2016.
* Robert I Kabacoff, _R in action_, 2/e (ch 19), Manning, 2015.
* Roger Peng, _Coursera: Exploratory Data Analysis_
* Brian Caffo, _Coursera: Developing Data Products_
* ggplot2 docs: http://docs.ggplot2.org/current/
* plotly: https://plot.ly/ggplot2/
